---
title: "homework3_sol"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## 1a

```{r}
y_A = c(12, 9, 12, 14, 13, 13, 15, 8, 15, 6)
y_B = c(11, 11, 10, 9, 9, 8, 7, 10, 6, 8, 8, 9, 7)

pr = rep(0, 50)
n0s = 1:length(pr)
for(n0 in n0s) {
  theta_A = rgamma(10000, 120 + sum(y_A), 10 + length(y_A))
  theta_B = rgamma(10000, 12 * n0 + sum(y_B), n0 + length(y_B))
  
  pr[n0] = sum(theta_B < theta_A) / length(theta_B)
}

plot(1:length(pr), pr, xlab = "n0", ylab = "MC estimate p(theta_B < theta_A)")
```

As the prior for B centered at 12 gets stronger, our posterior estimate that $p(\theta_B < \theta_A)$ gets lower.

## 1b

```{r}
y_A = c(12, 9, 12, 14, 13, 13, 15, 8, 15, 6)
y_B = c(11, 11, 10, 9, 9, 8, 7, 10, 6, 8, 8, 9, 7)

pr = rep(0, 50)
n0s = 1:length(pr)
for(n0 in n0s) {
  theta_A = rgamma(10000, 120 + sum(y_A), 10 + length(y_A))
  theta_B = rgamma(10000, 12 * n0 + sum(y_B), n0 + length(y_B))
  
  y_Ahat = rpois(length(theta_A), theta_A)
  y_Bhat = rpois(length(theta_B), theta_B)
  
  pr[n0] = sum(y_Bhat < y_Ahat) / length(y_Bhat)
}

plot(1:length(pr), pr, xlab = "n0", ylab = "MC estimate p(y_Bhat < y_Ahat)")

```

As the prior on $\theta_B$ around 12 gets stronger, the estimate of $p(\hat{y}_B < \hat{y}_A | y_B, y_A)$ gets lower.

## 1c

$\theta_B < \theta_A$ is the event that mice in group B accumulate cancer at a lower rate than mice in group A.

$y_B < y_A$ is the event that the actual number of mice in group B with cancer is less than the actual number of mice in group A with cancer.