---
title: "Homework 2"
output:
  html_document:
    df_print: paged
---

# Question 1

(a) I expect the average to be the same.

The variance for gamma(120, 10) will be lower.

(b) First we gotta compute our posteriors:

```{r cars}
yA = c(12, 9, 12, 14, 13, 13, 15, 8, 15, 6)
yB = c(11, 11, 10, 9, 9, 8, 7, 10, 6, 8, 8, 9, 7)

alpha_A = 120
beta_A = 10

alpha_B = 12
beta_B = 1

alpha_A_posterior = alpha_A + sum(yA)
beta_A_posterior = beta_A + length(yA)

alpha_B_posterior = alpha_B + sum(yB)
beta_B_posterior = beta_B + length(yB)
```

Distributions:
$$p(\theta_A | y_A) = \text{gamma}(11.9, .593)$$
$$p(\theta_B | y_B) = \text{gamma}(8.93, 0.638)$$

Means and variances:
```
print(paste0("Posterior mean of theta_A ", alpha_A_posterior / beta_A_posterior))
print(paste0("Posterior variance of theta_A ", alpha_A_posterior / beta_A_posterior^2))
print(paste0("Posterior mean of theta_B ", alpha_B_posterior / beta_B_posterior))
print(paste0("Posterior variance of theta_B ", alpha_B_posterior / beta_B_posterior^2))

#curve(dgamma(x, alpha_Ap, beta_A_posterior), 0, 20, col = 'blue')
#curve(dgamma(x, alpha_Bp, beta_B_posterior), 0, 20, add = TRUE, col = 'red')
#title('Posterior of theta_A in blue, theta_B in red')
```

Quantiles:
```{r}
print(paste0("Posterior 95% quantile for theta_A is [", qgamma(0.025, alpha_A_posterior, beta_A_posterior), ", ", qgamma(0.975, alpha_A_posterior, beta_A_posterior), "]"))
print(paste0("Posterior 95% quantile for theta_A is [", qgamma(0.025, alpha_B_posterior, beta_B_posterior), ", ", qgamma(0.975, alpha_B_posterior, beta_B_posterior), "]"))
```

(c) Write a little loop that computes the posterior of theta_B for each value of n0:

```{r}
means = rep(0, 50)
for (n0 in 1:50) {
  alpha_B_posterior = 12 * n0 + sum(yB)
  beta_B_posterior = n0 + length(yB)
  means[n0] = alpha_B_posterior / beta_B_posterior
}
plot(1:50, means, xlab = "n0", ylab = "posterior mean of theta_B")
```

(d) From the original question, we expect that the cancer rates of the two rats would be positively correlated.

We might want to use a prior that implies $\theta_A$ and $\theta_B$ are positively correlated. Assuming that $\theta_A$ and $\theta_B$ are independent does not help us really.

# Question 2

(a) 

- By the mean formula for gamma distribution, the mean rates are $\frac{3}{2000}$ and $\frac{7}{1000}$.

- Consider the variance of the prior distributions, $\frac{3}{2000^2}$ and $\frac{7}{1000^2}$, respectively.
Therefore, the second expert is more confident. 

(b)
```{r}
gamma_fun1 <- function(l) {2000^3/gamma(3) * l^(3-1) * exp(-2000 * l)}
gamma_fun2 <- function(l) {1000^7/gamma(7) * l^(7-1) * exp(-1000 * l)}

curve(0.5 * 2000^3/gamma(3) * l^(3-1) * exp(-2000 * l) 
              + 0.5 * 1000^7/gamma(7) * l^(7-1) * exp(-1000 * l),
      from = 0, to = 0.03, xname = 'l', xlab = "lambda", ylab = "density")
```

(c)

* 
\begin{align*}
posterior & \propto e^{-1767 * \lambda} * \frac{(1767 * \lambda) ^ 8} {8!} * (0.5 * \frac{2000^3}{\Gamma(3)} \lambda^{3-1} e^{-2000 * \lambda} + 0.5 * \frac{1000^7}{\Gamma(7)} \lambda^{7-1} e^{-1000 * \lambda}) \\
& \propto e^{-1767 * \lambda} *  \lambda ^ 8* ( \frac{2000^3}{\Gamma(3)} \lambda^{2} e^{-2000 * \lambda} +  \frac{1000^7}{\Gamma(7)} \lambda^{6} e^{-1000 * \lambda}) 
\end{align*}

(d)

* \begin{align*}
K &= \int^{\infty}_{0}{e^{-1767 * \lambda} * \frac{(1767 * \lambda) ^ 8} {8!} * (0.5 * \frac{2000^3}{\Gamma(3)} \lambda^{3-1} e^{-2000 * \lambda} + 0.5 * \frac{1000^7}{\Gamma(7)} \lambda^{7-1} e^{-1000 * \lambda})}d\lambda\\
&= \int^{\infty}_{0}0.5 * \frac{2000^3}{\Gamma(3)} * \frac{1767^8}{8!} \lambda^{10}e^{-3767 * \lambda}d\lambda + \int^{\infty}_{0}0.5 * \frac{1000^7}{\Gamma(7)} * \frac{1767^8}{8!} \lambda^{14}e^{-2767 * \lambda}d\lambda \\
&=  0.5 * \frac{2000^3}{\Gamma(3)} * \frac{1767^8}{8!} \int^{\infty}_{0}\lambda^{10}e^{-3767 * \lambda}d\lambda + 0.5 * \frac{1000^7}{\Gamma(7)} * \frac{1767^8}{8!}* \int^{\infty}_{0}\lambda^{14}e^{-2767 * \lambda}d\lambda\\
&= 0.5 * \frac{2000^3}{\Gamma(3)} * \frac{1767^8}{8!} * \frac{\Gamma(11)}{3767^{11}} + 0.5 * \frac{1000^7}{\Gamma(7)} * \frac{1767^8}{8!} * \frac{\Gamma(15)}{2767^{15}}
\end{align*}

* It becomes a mixture of Gamma(11, 3767) and Gamma(15, 2767)
$$posterior = \frac{1767^8 * 0.5 * 2000^3}{8!K \Gamma(3)}\lambda^{10}e^{-3767 \lambda} + \frac{1767^8 * 0.5 * 1000^7}{8!K \Gamma(7)}\lambda^{14}e^{-2767 \lambda}$$

(e)

```{r}
K = 0.5 * 2000^3/gamma(3) * 1767^8/factorial(8) * gamma(11)/3767^11 +
  0.5 * 1000^7/gamma(7) * 1767^8 / factorial(8) * gamma(15)/2767^15

const1 = 1767^8 * 0.5 * 2000^3/factorial(8)/K/gamma(3)
const2 = 1767^8 * 0.5 * 1000^7/factorial(8)/K/gamma(7)

curve(const1 * l^10 * exp(-3767 * l) + const2 * l^14 * exp(-2767 * l),
      from = 0, to = 0.03, xname = "l", 
      xlab = "lambda", ylab = "density")

# two rates
abline(v = 3/2000, col = 'blue')
abline(v = 7/1000, col = 'blue')

# mle
abline(v = 8/1767, col = 'red')
```

